The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-12-29 doc/ResponseLetter-R1.tex

\section*{}
\noindent To the Ed., AE, \& Editorial Board
\bigskip

\noindent Please consider this revision of our manuscript (MS \#60734), titled "The demographic costs of sexually antagonistic selection in partially selfing populations" for publication in \textit{The American Naturalist}. 
\bigskip

\noindent First of all, we want to thank the Ed., AE, and EiC for their thoughtful and forthright responses (and apologies) regarding the editorial delay on our manuscript. We understand completely that pandemic-related personal issues have been and continue to be extremely disruptive in these difficult times, and that editorial duties for our journals is a volunteer service. To our AE, Dr.~Orive, we sincerely hope that you and your family are well, and we truly appreciate the time and effort you took to evaluate our paper when you clearly had more imporant issues to deal with.
\bigskip

\noindent We received very helpful and constructive comments on our first submission and have now made a variety of major revisions to the MS in order to address them. First, we have made significant changes to our introduction to develop conceptual links between our model and previous work on evolutionarily explicit structured models and empirical research on sexually antagonistic selection. Second, we have tried to lessen the burden on non-theoretical readers by shifting development of the general model to the Supplementary Material, by following R2's many helpful suggestions regarding the presentation of the 2-stage model, and by including a new explanatory figure to help clarify the genetic and demographic outcomes from our model. Third, we have revised several figures to show population intrinsic growth rates ($\lambda$) rather than extinction boundaries. Finally, we have addressed the impacts of density dependence by ({\itshape i}) developing a new version of our model which includes simple density-dependence and reanalyzing the {\itshape M. guttatus} example; and ({\itshape ii}) expanding our discussion of density-dependence to address these additions. Overall, we feel that these various additions have greatly improved the paper. 
\bigskip

@@

\begin{quote}
	{\itshape This is a very good and totally valid point. To be honest, we believe that a separate paper is necessary to deal adequately with the many consequences of density dependence for our model predictions. However, we also agree that sidestepping this issue completely isn't good enough. In the end, we have made two main changes to address density dependence. First, have added a new {\bf Appendix B} where we briefly explain how density dependence can be incorporated into the model, and reanalyze our {\itshape M. guttatus} example with a simple DD version of the model. Second, we have expanded our methods, results, and discussion to more fully address the implications of DD, with reference to our modified figures and the new Appendix ({\bf L.150--156; 478--481; 564--585}).}
\end{quote}

@@

Related to that, and without doing more modeling, one quantity of interest you can present is the actual growth rate, and not just the region of viability/inviability. In a world where we acknowledge density dependence as a possibility, how far above or below one growth rate is can be informative about whether you actually expect extinction or depressed densities.

\begin{quote}
	{\itshape This is great suggestion. We have now included a new illustrative figure ({\bf fig.~1}), and revised versions of figs. 1 \& 4 (now {\bf figs.~S1 \& 5}) to show heatmaps of absolute values for predicted $\lambda$ instead of the extinction boundaries. However, one unfortunate consequence of presenting $\lambda$ instead of the extinction boundaries is that it is no longer possible to illustrate multiple values of maximum fertility ($f$) on each figure panel. In particular, our old fig. 1 now requires 18 panels to illustrate $\lambda$ for the same set of parameter combinations. We felt that this resulted in an unnecessarily complicated figure, especially since fig. 1's main purpose was to illustrate the general behaviour of the model for different fertility values. For simplicity, we have decided to retain our original version of fig.1 (now {\bf fig. 2}) in the main text and include the modified version showing $\lambda$ as a supplementary figure (now {\bf fig.~S1}); but will swap this figure if the editor/reviewers feel differently. However, we agree that the {\itshape M. guttatus} example was greatly improved by showing $\lambda$, and we now present the modified version of fig. 4 (now {\bf fig. 5}) in the main text}.
\end{quote}

@@

\noindent xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

\section*{Responses to Associate Editor Maria Orive}

Both reviewers found the work interesting and potentially to be of strong interest to the readership of AmNat, and I agree. However, Reviewer 1 brings up an important point regarding the framing of the work and placing it within a broader context of the literature on evolutionarily explicit structured models, which I think is important for the authors to do in order to make the paper accessible to and of interest for the overall readership of AmNat. There are other points that the reviewers bring up that are important to consider, but I think this one needs to be addressed carefully.

\begin{quote}
	{\itshape We agree that in our efforts to appeal to both demographers and evolutionary biologists studying sexual conflict, we ended up insufficiently discussing the relevant literature on both counts (but especially regarding evolutionarily explicit structured models). We have made a variety of changes to our introduction to improve this ({\bf L.45--51, L.69--71, L.76--85}). See also our specific responses to Reviewer 1. }
\end{quote}

@@

Reviewer 2 has some very nice suggestions regarding ideas for simplifying and clarifying the model description.

\begin{quote}
	{\itshape Agreed. Please see our sepecific responses to Reviewer 2 below.}
\end{quote}

In addition to the clarifications/edits and other points raised by the reviewers, I include a handful of minor edits/typos here of my own:
\bigskip

\noindent (1) Abstract, perhaps add a very general and short explanation of the term sexually antagonistic polymorphism.
\begin{quote}
	{\itshape We have added a brief general explanation of sexually antagonistic selection, which we felt was lacking in the first draft, and also allows us to more clearly link SA polymorphism with its' demographic consequences ({\bf L.2--3, L.6--7}).}
\end{quote}

@@

\begin{quote}
	{\itshape Thank you again for your efforts and consideration in spite of these challenges. We really appreciate it.}
\end{quote}

\noindent xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

\subsection*{Reviewer \#1:}

The manuscript reports the derivation, analysis, and example application of an extended frequency-dependent matrix model to explore links between sexual antagonistic selection, population persistence, and the maintenance of heritable variation. The paper is very well written and is straightforward to follow. The derivations are correct, although I did not try to derive all the equations in the appendices. My one comment is how might environmental stochasticity in the rate of selfing and outcrossing impact general conclusions? 
\begin{quote}
	{\itshape We now briefly discuss this possibility in the Discussion ({\bf L.605--611}).}
\end{quote}

@@

The manuscript adds to a growing and important literature where eco-evolutionary feedbacks are examined in structured models, some of which incorporate evolution explicitly. It also contributes nicely to a body of empirical and theoretical work on sexually antagonistic selection.

\begin{quote}
	{\itshape Thank you for these encouraging words.}
\end{quote}

As written the manuscript will not appeal to the broad readership of the American Naturalist because the weakness of the manuscript is the author makes little attempt to fit the work into the broader relevant literature. Instead, the paper has been written for a small audience of experts interested in the demographic consequences of sexually antagonistic selection. As written, the work is suited to a journal such as Journal of Theoretical Biology. This is disappointing, and the authors do themselves no favors by writing for such a niche audience. The manuscript could be well-suited to the American Naturalist if the authors were to engage with the broader literature. This can be done by extending the introduction to link this work to the wider literature on evolutionarily explicit structured models, and to the large empirical literature on sexually antagonistic selection. If they are able to do this appropriately, I believe this paper would make a strong contribution to the American Naturalist.

\begin{quote}
	{\itshape This is a really important comment and we thank you for pointing out this major shortcoming. One of our main goals when crafting the first draft of the manuscript was to find a middle ground that would appeal to both demographers and people studying sexual conflict. It is clear that we didn't quite achieve that goal. Rather, by making compromises in the introduction we ended up lacking scholarship on both sides. To address your main concern, we have extended the Introduction to more directly connect with the broader literature on evolutionarily explicit structured models (see {\bf L.76--85}) and now also provide a more thorough link with the empirical sexually antagonistic selection literature (see {\bf L.45--52}). We hope these changes succeed in broadening the appeal of our manuscript to the wider AmNat readership.}
\end{quote}

@@

\subsection*{Reviewer \#2:}

I have read, with great interest, this manuscript analyzing the effects of sexually antagonistic selection in hermaphrodites at both the population genetics and population dynamics levels. As a representative (I guess) of the general readership of AmNat and also as a population dynamics specialist, I appreciated the description of the model, its parametrization, the way it is constructed and analyzed. Not being a population genetics specialist, it is harder for me to comment on that side of the interpretations, apart from generic comments and in the light of the model’s characteristics. Overall, my take is that this paper is interesting, well written, the model well built and the ideas quite novel; and that it therefore corresponds to what readers except to find in AmNat both in terms of content and quality.

\begin{quote}
	{\itshape Thank you for the encouraging assessment.}
\end{quote}

\noindent I have one main comment:

\noindent-- Discussion: Conclusions are drawn before limitations of the model are mentioned. However, it seems to me that it would be very important to discuss the effect of those limitations on the results themselves and therefore discuss the generality of the results. For instance, what is the potential effect of female dominance in the marriage function on the results and in particular on conclusion of lines 477-480.

\begin{quote}
	{\itshape While we feel that there is strong precedence for discussing the key results of a study before diving into the various caveats that come with the analysis, the point is well taken that we need to clearly direct readers to the places in our paper where these limitations are discussed. In our Methods section, we are now careful to point readers to the various Appendixes, Supplements, and places in the Discussion where these limitations are dealt with (e.g., see {\bf L.101--103, 114--119, 151--157, 215--216, fig.~2 caption}), and we have added more explicit discussion of several of our key assumptions in the discussion ({\bf e.g., L.565--586; 606--612}). We discuss both the causes and potential effects of incorporating male-limitation of reproductive success on {\bf L.587--599}}.
\end{quote}

@@

\noindent Apart from that, I have only minor comments, that I list below (roughly in decreasing order of importance):
\bigskip

\noindent -- Model: lines 114-195
\begin{itemize}
	\item The matrices H are mentioned before they are introduced. (Therefore, the parenthesis in lines 123-124 cannot be understood by the reader). And then are only defined in the following section	
	\begin{itemize}
		\item It would make sense to reorder the model presentation by presenting the building blocks (arguably U, F and H matrices, but I'll come back to H in the next point), what they do, what they are for; then provide some properties (F independent from selfing/crossing, H from stage, etc ...) and then go into the details of their construction
	\end{itemize}
	\item The main issue I have here, is that H is introduced, but then disappears from the output (see equations 12 and 13 for instance that use H but it does not appear, or the general output eq 9)
	\begin{itemize}
		\item This suggests combining F and H directly into an object.
	\end{itemize}
\end{itemize}

\begin{itemize}
	\item More generally, the matrix formulations of the model may be a bit hard to read for the general reader of AmNat. Getting rid of H (or to the contrary, using it all the way, thus simplifying eq 12 and 13 for instance) and not giving a name to matrix W (which is just a bunch of 0s 1s and 1/2s) are steps that can make it easier for the reader to follow.
\begin{quote}
	{\itshape \ul{Response to the above 5 points}: As per the reviewers (and the editors) suggestion, we have moved a lot of the derivations into the supplementary materials, and now no longer use H or W in the main text. In doing so, we hope to shift the focus from mathematical derivations, which are probably unhelpful to many readers, towards explaining the end result instead. However, we have made the requested changes to clarify the derivation in the Supplement. }
\end{quote}

	\item It is quite confusing that HX is written as a constant (eq 7) whilst it depends on the frequencies and that HS is written as dependent (eq6) whilst it is constant
	\begin{quote}
	{\itshape That's a mistake on our part, apologies. We have corrected this in {\bf Online Supplement 1}, where it now appears.}
\end{quote}
	\item As the building blocks are given expressions (eq 10 and 11 for instance) it would be nice to mention the fact that the elements in these will be characterize later (otherwise the reader thinks she/he should already know what the expression of, say, UsAA is)
\begin{quote}
	{\itshape As suggested by this reviewer and by the editor, we have now moved up the two stage example, so that we show the reader what these matrices can look like immediately after introducing them. Again shifting the focus from mathematical derivations to an explicit example, thereby hopefully improving the manuscripts' accessibility/readability.  }
\end{quote}

@@

\end{itemize}

\noindent -- Line 19: I find it weird to write that the "process of adaptation can be impeded by ...changing environmental conditions".

\begin{itemize}
	\item One organism can be adapted to an environment (as in a series of environmental parameters) but also to changes in that environment (plasticity, genetic variance etc ...). It is actually from this consideration that a lot of studies on adaptation and adaptive dynamics.
	\item I think this sentence should be rewritten to make it clear that it is not changes in environmental conditions per se that impede the adaptive process but some of their characteristics (speed, range etc.)
\end{itemize}

\begin{quote}
	{\itshape Very good points, that was awkward wording on our part. We have rephrased to more clearly explain that it is primarily the speed/magnitude of environmental change that we are referring to ({\bf L.23}). We have also included a new citation of a timely AmNat paper that explores some of these issues explicitly.}
\end{quote}

@@

\noindent -- Line 23: The link between genetic constraints and genetic trade-offs should be clarified. Or rather a definition of each term provided, for the reader to understand this sentence. I suspect that, for many a reader of AmNat, genetic constraints and genetic trade-offs are equivalent concepts.

\begin{quote}
	{\itshape Agreed. We have rephrased to avoid the term "trade-off" which tends to over-simplify the concept in our opinion ({\bf L.28}).}
\end{quote}

@@

\noindent -- Line 30: "fitness trade-offs ... provide ...mechanism for the maintenance of genetic variation". This requires a few words of explanation. Is that a general statement about (fitness) trade-offs? or is it actually particular to SA trade-offs or to trade-offs between fixed heterogeneity components?
\begin{itemize}
	\item For instance, a fitness trade off between fertility at 3 years and fertility at 6 years will have no positive effect on genetic variance. To the contrary, it may lead the organism to evolve an optimal strategy accounting for overall fitness and the trade-offs, vis-à-vis investments in reproduction at those different stages in the life cycle. And as such this may lead to the selection of individuals with such a strategy and therefore to low genetic variance.
\end{itemize}

\begin{quote}
	{\itshape We have rephrased the first sentence of this paragraph to be more specific ({\bf L.30}), which we hope also addresses this comment.}
\end{quote}

@@

\noindent -- Line 264: this is an important point that is a bit unclear here, according to me. Could you clarify that what you call demographic rates are the transitions in the overall model and so include both demography and genetics? Clarify also that the model is not density dependent because it only depends on frequencies ... and maybe relate it to a general equation where A does not depend on n per se but on the q frequencies.

\begin{quote}
	{\itshape \textcolor{red}{Lotte, double check here?} We have added a few lines to discuss frequency vs density dependence when we first introduce the population projection matrix in equation 3 ({\bf L.151--157}). In addition, we have tried to clarify to improve this section by discussing that the survival and fertility rates ("demographic rates") are density-indepedent, and that the allele frequency dynamics make the model frequency dependent.}
\end{quote}

@@

- Line 35: SA selection arises also from non-genetic trade-offs, which makes the formulation of that sentence a bit weird.

\begin{quote}
	{\itshape We have rephrased this sentence to clarify and be more consistent with our language earlier in the preceding paragraphs ({\bf L.40--42}).}
\end{quote}

@@

- Line 42: Define analogous here

\begin{quote}
	{\itshape This sentence has been rephrased.}
\end{quote}

- Line 54: maybe a few words to remind the reader about what dioecious and hermaphrodite organisms are.

\begin{quote}
	{\itshape We have added a sentence earlier in this paragraph where we remind readers what dioecy is ({\bf L.57}). Hermaphroditism is defined earlier in the paragraph, so an additional reference seemed redundant.}
\end{quote}

@@

- Line 233: f and f’ are called "adult fertilities". However, they correspond to that rate only for certain specific genotypes. Maybe clarify that by calling them "adult maximum fertilities" ?

\begin{quote}
	{\itshape Good point, fixed ({\bf L.193}).}
\end{quote}

- Results: it took me a while to figure out Fig 1 (mainly because I was looking at it in B and W). I am just wondering whether it would be easier to read with only one f per chart. Maybe as a separate chart that illustrates the general shape of the parameter space for polym * extinction for only one C, one h and one f... then followed by these 6 graphs as they are now.

\begin{quote}
	{\itshape Thank you for this suggestion. To address this issue, as well as the Editor's request that we present population growth rates, we now start with an illustrative figure that only has one $f$, and that shows separately what happens to the allele frequencies (Population Genetic Outcome) and to the population growth rate (Demographic Outcome), see new {\bf fig. 1}. We hope that this will make it easier to interpret the old fig. 1, which is now {\bf fig. 2}. We have also included a new supplementary figure, {\bf fig. S1}, with detailed heatmaps for each of the parameter combinations presented in this figure (i.e., one $f$ per panel). However, we felt that this revised figure was to large and complicated to present in the main text (18 vs. 6 panels).}
\end{quote}

\noindent-- Maybe mention once, in a few words, what is meant by "segregate as a diallelic locus".
\begin{quote}
	{\itshape We have added a brief parenthesis to clarify ({\bf L.447--448}).}
\end{quote}

@@

\noindent-- Fig 4 : they are supposed to be dashed lines and dotted lines but all I see are dashed lines with various dashes. I found 4B in particular, hard to understand at first sight.

\begin{quote}
	{\itshape This figure now presents four separate panels with heatmaps of $\lambda$, each with one set of invasion conditions ({\bf fig.~5}). We hope this eases interpretation.}
\end{quote}
\bigskip

@@

\noindent-- The 0s representing 0 matrices in eq 5 are not bold
\begin{quote}
	{\itshape Fixed}
\end{quote}

\noindent-- In same equation 1w is not defined
\begin{quote}
	{\itshape Fixed}
\end{quote}

\noindent-- Line 142, 2 "the"s
\begin{quote}
	{\itshape Fixed}
\end{quote}

\noindent-- Line 553 : not a typo, but the "was that how" in the middle of the sentence makes it hard to read
\begin{quote}
	{\itshape We have re-written this sentence to avoid this awkward phrasing ({\bf L.600--601}).}
\end{quote}

### 2021-12-29 doc/SA-Hermaphrodites-wDemography-R1.tex
2nd pass edits
2nd complete pass at edits in response to reviewer comments.

% Alternative opening sentence to the abstract:sentence
Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying parameter conditions under which SA polymorphism is maintained {\itshape and} the population growth rate remains positive. \textcolor{blue}{A sexually antagonistic polymorphism is a polymorphism of two alleles with antagonistic effects on the two sexes, that is, one of the two coexisting allele confers a fitness benefit to females and reduces male fitness, and the other allele confers a fitness benefit to males and reduces female fitness. } Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

% GOOD STUFF HERE... KEEPING THIS IN CASE WE WANT TO WORK MORE OF THIS INTO THE ABSTRACT...In this paper we developed a stage-structured mendelian matrix model that accommodates selection through male and female sex-functions in populations of simultaneous hermaphrodites with mixed mating systems. Using the model, we jointly analyzed the evolutionary and demographic consequences of SA selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focused our analyses on identifying the parameter conditions under which SA polymorphisms are maintained by balancing selection {\itshape and} the population is also demographically viable (i.e., has a positive intrinsic growth rate). We also looked at the effects of inbreeding depression manifesting in different life-history stages, exploring the fitness consequences of early- vs.~late-acting inbreeding depression on the model predictions. Finally, we gave an illustrative example of the potential for SA polymorphism in real populations by parameterizing our model with empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).

@@

To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{there is a negative correlation between different fitness components} \citep{CharlesworthHughes2000, ConnallonHall2018}.

\textcolor{blue}{Genetic constraints on adaptation arising from conflicting selection and gene flow between different classes of individuals} have particularly interesting evolutionary consequences \citep{ConnallonDebarreLi2018}. On the one hand, they provide an effective mechanism for the maintenance of genetic variation, which can increase a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that arises when beneficial alleles for one sex are deleterious when expressed in the other. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Likewise, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness, indicating SA fitness effects for multilocus genotypes (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}). Moreover, polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}.}

@@

\textcolor{blue}{While significant efforts have been made to understand when balanced SA polymophisms can be maintained and to detect SA genetic variation from empirical data, much less consideration has been given to the demographic consequences of the maladaptation caused by SA selection}. Most population genetic models, including models of SA selection, model the relative fitness of genotypes in populations of constant (and often infinite) size, and do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{that link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). Over the past 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. In particular, theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic changes accompanying allele frequency changes due to selection. Extensions of these theories to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects} all highlight the sometimes counterintuitive demographic consequences of adaptive evolution that are missed by classical population genetics models. }

%\textcolor{blue}{One of the earliest such models \citep{roughgarden1971density} combined a population genetic model with a density-dependent population model with no intraspecific demographic structure. Charlesworth then combined an age-structured population model with population genetics in a series of papers, consolidated in \cite{Charlesworth_1994}. \cite{orive1995senescence} extended this work to study life-history evolution in organisms with both clonal and sexual reproduction (see also \cite{orive2001somatic}).  \cite{ronce2000kin} added spatial structure and stochasticity to age-structured genetic models to model kin selection, a body of work consolidated in \cite{rousset2004genetic}. Of course there are many other approaches to studying eco-evolutionary dynamics, e.g. combining quantitative genetics models with demographic structure (e.g. \cite{coulson2006putting,barfield2011evolution,childs2016evolution,orive2017effects}) or adaptive dynamics approaches (e.g. \cite{dieckmann2002adaptive,durinx2008adaptive,ten2020habitat}). For a comprehensive review of the eco-evolutionary literature, see \cite{fussmann2007eco}, or more recently, \cite{govaert2019eco}.}

@@

Simultaneous hermaphrodites can transmit genes to the next generation via both sperm/pollen and eggs/ovules, and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from another individual in the population, while paternal outcrossing involves exporting male gametes to another individual and fertilizing their ovules. Self-fertilization is achieved when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female function, we denote matrices or vectors relating to the male sex function with a prime. \textcolor{blue}{Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen, for reasons of analytic tractability. We present a more general model of genotype-specific self-fertilization rates (after \citealt{JordanConnallon2014}) in the Online Supplement, and briefly discuss implications of this approach in the Discussion.}

@@

\textcolor{blue}{To demonstrate how to construct and analyze a genotype $\times$ stage-classified model, we will now construct an example model for a hypothetical species with intralocus sexual conflict via the two sex functions. }For the sake of simplicity, we assume our hypothetical species has a life cycle with only two stages: juveniles and adults (i.e., $\omega = 2$), and that only adults are reproductively active. Suppose that there is a genetic trade-off between the sex-functions at a single diallelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect. Following convention, we parameterize the fertility component of fitness for each genotype through each sex function, $w_{i}$ and $w^{\prime}_{i}$, to be bounded by $[0,1]$, with dominance and selection coefficients $h_f, s_f$ and $h_m, s_m$ determining the decrease in fertility through each sex function relative to the most fit genotype ($AA$ has highest female fertility, $aa$ the highest male fertility; see Table \ref{tab:Fitness}).

@@

Iterating the projection matrix, Eq(\ref{eq:Atilde}) with the above demographic matrices, given an initial population state vector, allows numerical simulation of the eco-evolutionary dynamics for selection operating on any of the sex-function, or stage-specific demographic parameters. As we outline below, we use numerical techniques together with mathematical analyses to study the conditions for the maintenance of SA polymorphisms, and the demographic fate of the populations (i.e., positive growth, or extinction).

>>>>

% Alternative opening sentence to the abstract:sentence

\textcolor{blue}{When selection differs between the sexes, genes expressed by both males and females can experience sexually antagonistic (SA) selection, where beneficial alleles for one sex are deleterious for the other.} Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of such SA genetic polymorphisms, but these models have rarely considered the demographic consequences \textcolor{blue}{ of the maladaptation imposed by coexisting alleles with deleterious fitness effects in each sex.} In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying when SA polymorphisms are maintained by balancing selection {\itshape and} the population growth rate remains positive. Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations can sustain large regions of viable SA polymorphic space.

%\textcolor{blue}{A sexually antagonistic polymorphism is a polymorphism of two alleles with antagonistic effects on the two sexes, that is, one of the two coexisting allele confers a fitness benefit to females and reduces male fitness, and the other allele confers a fitness benefit to males and reduces female fitness. }

@@

@@

To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}.

\textcolor{blue}{Genetic constraints on adaptation arising from conflicting selection and gene flow} have particularly interesting evolutionary consequences \citep{ConnallonDebarreLi2018}. On one hand, they provide an effective mechanism for the maintenance of genetic variation, which can increase a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Likewise, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and negative correlationpolygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}.}

@@

\textcolor{blue}{While significant efforts have been made to understand when balanced SA polymophisms can be maintained and to detect SA genetic variation from empirical data, much less consideration has been given to the demographic consequences of the maladaptation caused by SA selection}. Most population genetic models, including models of SA selection, model the relative fitness of genotypes in populations of constant (and often infinite) size, and do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{which link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). Over the past 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. In particular, theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic changes accompanying allele frequency changes due to selection. Extensions of these theories to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects} all highlight that a fuller understanding of the genetic basis of fitness variation, and the processes that maintain it, requires an explicit treatment of the demographic consequences of evolutionary change.}

@@

Simultaneous hermaphrodites can transmit genes to the next generation via both sperm/pollen and eggs/ovules, and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from other individuals in the population, while paternal outcrossing involves exporting male gametes to other individuals and fertilizing their ovules. Self-fertilization occurs when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female reproductive functions, we denote matrices or vectors relating to the male sex function with a prime. \textcolor{blue}{Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen, for reasons of analytic tractability. We present a more general model of genotype-specific self-fertilization rates in the Online Supplement, and briefly discuss implications of this approach in the Discussion.}

@@

\textcolor{blue}{To demonstrate how to construct and analyze a genotype $\times$ stage-classified model, we will now construct an example model for a hypothetical hermaphrodite species with intralocus sexual conflict via the two sex functions. }For the sake of simplicity, we assume our hypothetical species has a life cycle with only two stages: juveniles and adults (i.e., $\omega = 2$), and that only adults are reproductively active. Suppose that there is a genetic trade-off between the sex-functions at a single diallelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect. Following convention, we parameterize the fertility component of fitness for each genotype through each sex function, $w_{i}$ and $w^{\prime}_{i}$, to be bounded by $[0,1]$, with dominance and selection coefficients $h_f, s_f$ and $h_m, s_m$ determining the decrease in fertility through each sex function relative to the most fit genotype ($AA$ has highest female fertility, $aa$ the highest male fertility; see Table \ref{tab:Fitness}).

@@

Iterating the projection matrix, Eq(\ref{eq:Atilde}), with the above demographic matrices, given an initial population state vector, allows numerical simulation of the eco-evolutionary dynamics for selection operating on any of the stage- or sex-function-specific demographic parameters. As we outline below, we use numerical techniques together with mathematical analyses to study the conditions for the maintenance of SA polymorphisms, and the demographic fate of the populations (i.e., positive growth, or extinction).
